---
title: "Small Area Estimation of Poverty in Somalia : A Fay Herriot Model Approach"
author: "Ifeanyi Edochie"
date: "2024-06-20"
output: 
  bookdown::word_document2:
    reference_docx: report_style.docx
fontsize: 10pt
always_allow_html: true
bibliography: bibliography.bib
nocite: | 
  @CARRAO2016108, @nelson2019suite, @nhess-20-695-2020
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# BiocManager::install("kable", update = FALSE, validate = FALSE)

pacman::p_load(here, ggplot2, sf, sfnetworks, kableExtra, dplyr, data.table, 
               viridis, paletteer, ggthemes, gridExtra, grid, flextable,
               showtext, extrafont, bookdown, officer, ggExtra)

sf::sf_use_s2(FALSE)


source(here("analysis/01_fayherriotmodel.R"))


median_dist_cv <- round(median(spatial_dt$CV, na.rm = TRUE)/100, 2)
mean_dist_cv <- round(mean(spatial_dt$CV, na.rm = TRUE)/100, 2)

median_region_cv <- round(median(provpov_dt$CV/100, na.rm = TRUE), 3)
mean_region_cv <- round(mean(provpov_dt$CV/100, na.rm = TRUE), 3)

```
### Introduction

Household surveys are often representative at the national level or at the level of the first administrative division (region/state level). National Statistical Offices and government entities can benefit from poverty estimation at a higher level of resolution, such as the district or district level. This note describes the small area estimation (SAE) methodology implemented to estimate poverty rates in Somalia at the district level. SAE is a statistical method that can be used to improve the reliability of survey estimates by combining survey data with geographically comprehensive auxiliary data, such as census when available or geospatial, remotely sensed data. In Burundi, we show that SAE generates poverty estimates that are sufficiently precise to report at the district level instead of the regional level. This has the potential to improve the targeting and evaluation of interventions intended to achieve poverty reduction in the future. Ideally, SAE combines survey data with household-level data from a recent census. Countries often aim to collect census data every 10 years. However, many African countries take more years between consecutive censuses, and indeed Somalia’s last census was conducted in 1986. Therefore, this exercise relies on contemporaneous geospatial data derived from a variety of sources. [@battese1988error] were the first to use geospatial satellite data in the context of crop production. Other studies [@georganos2019modelling], [@chi2022microestimates] have used satellite imagery to predict wealth indices. However, the use of geo-referenced data is less ideal than the traditional microdata obtained from household surveys or administrative datasets. [@van2023accurate] used satellite imagery to predict monetary poverty in Malawi and noted the less than ideal nature of geospatial data. [@CORRAL2025103377] also find tat remotely sensed data may not be ideal for poverty mapping due to the relatively lower predictive power.   

In this note, we present the approach that models poverty rates at the district level in Somalia using the model of [@fay1979estimates]. The area level model approach allows us to relate district level direct survey poverty rates to auxiliary variables (geospatial indicators) to estimate poverty rates in all districts within Somalia. [@seitz2019they] provides district level poverty rates in the Central Asia region using Fay Herriot modelling approach using auxiliary geospatial data. The World Bank has employed the SAE methodology extensively to estimate poverty and other socioeconomic indicators of interest at more granular levels and continues to produce these estimates in combination with other non-monetary measures of poverty. At this point, SAE has been applied in a wide variety of contexts across many developing countries. This note is subdivided as follows. In section 2, we present survey data (specifically the household consumption data) and why SAE is necessary for district level poverty estimation in Somalia. We also present the Fay-Herriot model as described by [@seitz2019they]. Section 3 describes the geospatial databases sourced and indicators created as well as the model selection process employed. Sections 4 and 5 describes the FH model results and the poverty maps for the country.  


### The Data

For Somalia, the 2022 Somalia Integrated Household Budget Survey (SIHBS) is representative at the regional level. The development of the SIHBS-22 sampling frame followed a stratified multi-stage probability cluster sample design. Urban and rural areas followed a three-stage stratified cluster sample design, while in nomadic areas the design was a two-stage stratified cluster sample design. The primary sampling units (PSUs) were selected with a probability proportional to the number of dwelling structures which constituted the sampling frame. The secondary sampling units (SSUs) for rural and urban areas were selected with a probability proportional to the number of listed households which constituted the frame. The ultimate sampling units (USUs) for rural, urban, and nomadic areas were randomly selected from listed households in the cluster.  

District level poverty rate estimates computed from this survey will be insufficiently precise and unreliable for publication. \@ref(tab:tab1) illustrates why it is necessary to use SAE to report poverty rates at more disaggregated geographic levels in Somalia. We use the mean coefficient of variation (CV) as a standardized measure of precision (i.e.the square root of the estimated mean square error divided by the poverty rate). Differing thresholds for mean or median CVs, often ranging from 0.1 to 0.3, have been applied by National Statistics Offices to determine if statistics are sufficiently reliable to report. The median and mean direct CVs in Somalia at the district level are approximately `r median_region_cv` and `r mean_region_cv`. While this is within the acceptable range of reliability for some countries, it is not considered reliable enough to publish by the Somalia National Statistics Office.


```{r, echo = FALSE}

specify_decimal <- function(x, k) trimws(format(round(x, k), nsmall=k))

descriptives_dt <- 
  data.table(tot_pop_million = specify_decimal(sum(geosurvey_dt$population_weight, na.rm = TRUE)/1e6, 1),
             pop_hhs_million = specify_decimal(sum(geosurvey_dt$wgt_adj2, na.rm = TRUE)/1e6, 1),
             svy_hhs = nrow(geosurvey_dt),
             povrate_ipl = 
               geosurvey_dt %>%
               summarise(povrate_ipl = weighted.mean(x = poor,
                                                     w = population_weight,
                                                     na.rm = TRUE)) %>%
               specify_decimal(3),
             recent_census = 1986 %>% as.integer(),
             region_count_sample = length(unique(geosurvey_dt$admin1Pcod[!is.na(geosurvey_dt$admin1Pcod)])),
             median_cv_region = 
               directpov_dt %>%
               summarize(median = median(directpov_dt$CV/100, na.rm = TRUE)) %>%
               specify_decimal(3),
             mean_cv_region = 
               directpov_dt %>%
               summarize(mean = weighted.mean(x = directpov_dt$CV/100, 
                                              w = directpov_dt$SampSize,
                                              na.rm = TRUE)) %>%
               specify_decimal(3),
             num_targets_pop = length(unique(spatial_dt$admin2Pcod[!is.na(spatial_dt$admin2Pcod)])),
             num_targets_sample = length(unique(geosurvey_dt$admin2Pcod[!is.na(geosurvey_dt$admin2Pcod)]))) %>%
  t() %>%
  data.frame()

colnames(descriptives_dt) <- "Estimate"

  data.table(Indicator = c("Population (in millions)", 
                           "Population Number of HHs (in millions)",
                           "Sample # of HHs",
                           "Poverty Rate (IPL)",
                           "Latest Census Year",
                           "Number of Regions",
                           "Region Median CV", 
                           "Region Mean CV",
                           "Number of Targets (Population)",
                           "Number of Targets (Sample)"),
             Estimate = descriptives_dt$Estimate) %>%
    flextable() %>%
    colformat_num(j = "Estimate", digits = 3) %>%
    colformat_num(j = "Estimate", i = c(3, 5, 6, 9, 10), digits = 0) %>%
    colformat_num(j = "Estimate", i = 5,  big.mark = "") %>%
    set_table_properties(width = 0.5, layout = "autofit") %>%
    fontsize(size = 10) %>%
    font(fontname = "Times New Roman", part = "all") %>%
    autofit() %>%
    theme_box() %>%
    set_caption(caption = "Descriptive Statistics",
                style = "Table Caption",
                autonum = run_autonum(seq_id = "tab", 
                                      bkm = "tab1"))
  
```

We utilize freely available geospatial data for this small area estimation exercise since the last census carried out is outdated (from 1986). The goal of the SAE exercise is to estimate more reliable district level poverty rates in Somalia by using a Fay Herriot model based on relating the target area direct estimate poverty rates and district level geospatial indicators. Given that any recent developments in Somalia might not be captured by its 15-year-old census, it would be difficult to make a case for area poverty rates estimated using the 1986 census particularly to guide current policy interventions. 


### The Methodology
[@corral2022guidelines] recommends implementing an area level Fay Herriot level model with geospatial indicators for poverty mapping. Imagine a finite population for Somalia, \( P \), that consists of \( N \)  households that are subdivided into \( D \) districts with sizes \( N_1, \ldots, N_D \). A random sample of households can be drawn from the \( d^{th} \) commune (i.e., \( n_1, \ldots, n_d \quad \text{s.t.}\quad \; n < N \). The Fay-Herriot (FH) model comprises of two levels. The first is a sample model which assumes a direct survey estimator: 

$$
\hat{\theta}_i^{Dir} = \theta_i  + e_i, \qquad\forall_i = 1,\ldots,D
$$
\(  \hat{\theta}_i^{Dir} \) is design unbiased for the small area parameter, \( \theta_i \) the population indicator of interest, in this case, the poverty rate each district, \( d_i \). We assume a sample error \( e_i \) with the usual independently and normally distribution properties. 

$$
e_i \sim N(0, \sigma_{e_i}^2)
$$

In the second level, a linking model is assumed to relate \( \theta_i \) to auxiliary variables \( x_i = (x_{i1}, \ldots, x_{ic})' \) via a linear regression. Both levels of the model together are presented as follows:

$$ \hat{\theta}_i^{Dir} = x_i^{\mathcal{T}}\beta + \mu_i + e_i; \quad \mu_i \sim N(0, \sigma_{\mu}^2); \quad \forall i = 1,\ldots,D 
$$

The empirical best linear unbiased estimators (EBLUP) \( \beta \) are computed with by weighted least squares regression. The EBLUP of \( \theta_i \) is obtained by substituting the variance parameter \( \sigma_{\mu}^2 \) with an estimate. The resulting estimator can then be written as: 

$$ \hat{\theta}_i^{FH} = x_i^{\mathcal{T}}\hat{\beta} + \mu_i $$

$$ \hat{\theta}_i^{FH} = \hat\gamma_i \theta_i^{\mathcal{T}} + (1 - \hat\gamma_i)x_i\hat{\beta} $$

The EBLUP/FH estimator can be understood as a weighted average of the direct estimator and a regression synthetic part. The estimated shrinkage factor \( \hat\gamma_i = \frac{\hat\sigma_{\mu}^2}{\hat\sigma_{\mu}^2 + \hat\sigma_{\varepsilon_i}^2} \) puts more weight on the direct estimator when the sampling variance is small and vice versa. Areas for which no direct estimation results are called out-of-sample domains. For those domains the prediction reduces the regression-synthetic component \( \hat{\theta}_{i,out}^{FH} = x_i^{\mathcal{T}}\hat{\beta} \) [@molinarao]. This model is then fit via a restricted maximum likelihood (REML) method. 

This method is widely used by the Small Area Income and Poverty Estimates (SAIPE) program of the US census bureau and has been thoroughly validated in [@corral2023poverty]. This approach improves the error efficiency rates over the direct estimates at the target area level. Inter-area unexplained heterogeneous area effects are accounted for within the model. Section 3.3 in [@corral2022guidelines] provides a full list of pros and cons of the Fay-Herriot modelling approach. 

For this small area estimation exercise, 48 of the 74 Somalia Districts are included in the SIHBS. As a result, these in-sample districts will benefit from the information available in the sample. It is sometimes the case with Fay Herriot modelling that districts with low sample sizes can result in all households from a specific sample district being all poor or not poor (${\theta}_i^{Dir}$ = 1 or 0) or only one Enumeration Area in a district is sampled. The common practice of sample variance smoothing [@you2012sampling; @yongmike2023] in the SAE literature is typically implemented to solve this problem. The variance smoothing approach of [@you2012sampling] applies a log linear model of the direct sampling variance{$\hat{V}_i$} as a function of the sampling size, $n_i$. 

$$ log(\hat{V}_i) = \hat\phi{_0} + \hat\phi{_1}n_i + \varepsilon_i, \quad i = 1, ..., D $$
Assuming $\hat\phi{_0}$ and $\hat\phi{_1}$ to be the simple OLS estimators for the regression coefficients $\phi{_0}$ and $\phi{_1}$. Applying the exponential of the equation produces the naive variance estimator [@dick1995] as follows: 

$$ \hat{V}_i = e^{\hat\phi{_0} + \hat\phi{_1}n_i}$$
[@rivbel2000] show that the naive estimator can underestimate sampling variance. They propose correction as follows: 

$$ \hat{V^{RB}_i} = \hat{V_i}e^\frac{\tau^2}{2} $$
since the naive variance estimator can be easily shown to overestimate sampling variance by a factor of $e^\frac{\tau^2}{2}$. For the purposes of our analysis, there are no districts without variances or extreme case poverty rates (${\theta}_i^{Dir}$ = 1 or 0). Consequently, we do not remove districts from the analysis. However, the National Statistical Office flagged predicted poverty rates in 3 areas as being too low. 2 of these 3 districts have low sample sizes and with only 1 and 2 primary sampling units as well. In the supplementary section, we re-estimate the Fay Herriot model without these areas and present the results. Below, we simply show the sample distribution of PSUs and households within districts and how this varies with the direct district poverty rates. 

```{r figa, echo = FALSE, fig.cap = "Joint Distribution of EAs and Households (District-Level)"}

p <- 
smoothcheck_dt %>%
  merge(spatial_dt[, c("Direct", "admin2Pcod")]) %>%
  ggplot() + 
  geom_point(aes(x = num_ea, y = num_hhs, color = Direct), size = 3, alpha = 0.5) + 
  scale_color_viridis(option = "magma", direction = -1) +
  xlab("Number of EAs in District") + 
  ylab("Number of Households in District") + 
  theme_minimal() + 
  theme(legend.position = "left") 

ggExtra::ggMarginal(p, type = "histogram")

```


### The Geospatial Data and Model Selection Process
The process leading up to model selection involves sourcing freely available geospatial indicators that might be correlated with household welfare and poverty. The geospatial features were sourced at native resolution and then zonal statistics were computed at the target area level (districts). \@ref(tab:tab5) shows all the geospatial features and the data sources employed.  

```{r, echo = FALSE}

# Create a data frame to represent your table
data.frame(
  Feature = c("Built-settlement extent area", 
              "Gridded Population & Density", 
              "Share of area planted by crop for banana, beans, cassava, maize, sesame seed, sorghum, sugar cane, temperature fruit,
              tropical fruit, vegetables", 
              "Production quantity for each crop for banana, beans, cassava, maize, sesame seed, sorghum, sugar cane, temperature
              fruit, tropical fruit, vegetables", 
              "% production as a total crop production for banana, beans, cassava, maize, sesame seed, sorghum, sugar cane,
              temperature fruit, tropical fruit, vegetables", 
              "Standardized precipitation evaporation index, 12 month", 
              "Drought exposure, Drought hazard, Drought risk index, Drought vulnerability", 
              "Drought hazard, risk for irrigated agricultural systems", 
              "Percent of area (with Vegetation Index below 40) for the Gu season (April - June)", 
              "Average travel time in nearest urban areas with 5000km, 20000km and 50000km"),
  Source = c("WorldPop Building Footprints", 
             "WorldPop Gridded Population Counts & Density", 
             "IFPRI Spatial Production Allocation Model (SPAM)", 
             "IFPRI Spatial Production Allocation Model (SPAM)", 
             "IFPRI Spatial Production Allocation Model (SPAM)", 
             "Global SPEI database, version 4.03", 
             "(Carrao et al. 2018)", 
             "Drought risk for rainfed, irrigated agric. systems aggregated as an average per polygon based on the data from
             (Meza et al. 2020)", 
             "STAR - Global Vegetation Health Products", 
             "Computed based on population data from WorldPop and accessibility data from (Nelson et al. 2019)"),
  Original_Data_Resolution = c("1km", "90m", "10km", "10km", "10km", "0.5 degrees", "0.5 degrees", "", "", ""),
  Year = c("2001 - 2020", "2020", "2009, 2017, 2020", "2009, 2017, 2020", "2009, 2017, 2020", "2020", "2000-2014", "2020",
           "2017-2022", "2019")) %>%
  flextable() %>%
  set_table_properties(width = 1, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() %>%
  set_caption(caption = "EBP Model (Regression Results)",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab5")) %>%
  theme_box()

```


We begin by transforming all indicators as necessary to minimize the risk of divergence in model parameter estimation. For indicator with values greater than 100, we take the natural logarithm. We have avoiding feature scaling to avoid excessive distortion or loss of information for the scaled variables.

The geospatial data listed under the previous header was used to construct candidate features, at the grid and target area level. In addition, we include regional dummy variables. In all, we created ~157 potential geospatial candidate indicators. Using all these features in the linear mixed model risks potentially leads to over-fitting the survey sample and generates poor out-of-sample estimations. 

Next, we employ a stepwise (both-ways method) selection approach which picks the most predictive set of indicators from the pool of candidate indicators. The both-ways method was used, enabling iterative testing of each variable's contribution by alternatively adding and removing variables based on statistical significance criteria at each step. This approach begins with an constant term and tests the inclusion of variables one-by-one; then it considers each for potential removal, thus optimizing the model's explanatory power while controlling for over-fitting. The both-ways method provides flexibility, more so than the forward or backward algorithm, to achieve an optimal balance of predictive power and model parsimony, ensuring the only variables with significant and robust relationships to the outcome are retained. 

### Fay Herriot Model Estimation Results
The final selected model includes covariates share of the population within a 2km grid as well as the share of total production that is tropical fruit production. The regression results are as follows: 

```{r, echo = FALSE}

fh_reportcoef_table(fhmodel_not) %>%
    setnames(new = c("Variables", "Coefficients", "Standard Error")) %>%
    mutate(Variables = case_when(
    Variables == "(Intercept)" ~ "Intercept",
    Variables == "prodmt_reg_shr_trof" ~ "Production share in Tropical Fruit Production",
    Variables == "share_pop_in_2km_grid" ~ "Average share of population within 2km grids",
    TRUE ~ Variables
  )) %>%
  flextable() %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() %>%
  set_caption(caption = "EBP Model (Regression Results)",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab2")) %>%
  theme_box()

norm_dt <- 
fh_normality_fit(fhmodel_not) %>%
  mutate(value = specify_decimal(value, 3)) %>%
  t() %>%
  as.data.frame() 

```

```{r, echo = FALSE}


colnames(norm_dt) <- norm_dt[1,]
norm_dt <- norm_dt[-1,]

norm_dt %>%
  as.data.frame() %>%
  flextable() %>%
  set_header_labels(rsq_marginal = "marginal",
                    rsq_conditional = "conditional",
                    epsilon_skewness = "skewness",
                    epsilon_kurtosis = "kurtosis",
                    random_skewness = "skewness",
                    random_kurtosis = "kurtosis") %>%
  add_header_row(values = c("Model R\u00B2", "(Error Term) \u03B5", "(Random Effect) \u03BC"), 
                 colwidths = c(2, 2, 2)) %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  theme_box() %>%
  autofit() %>%
  set_caption(caption = "Assessing Normality Assumptions",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab3"))

```


The regression coefficients have the expected sign. The share of the population within the 2km grid is decreasing in poverty i.e. districts with more densely populated grids appear to be more well off (i.e. less poor). This may indicate that higher density areas attract individuals with more income or earning potential. 

Several assumptions are made in this model which needed to be verified. The Fay Herriot model \( R^2 \) equals `r as.numeric(norm_dt$rsq_conditional)*100` with an adjusted \( R^2 \) of `r as.numeric(norm_dt$rsq_marginal)*100` which is typical for the FH model particularly with only 48 in-sample districts used in the regression of only geospatial features. We assume independent normal distributions for the area effects as well as error terms. The table shows the skewness and kurtosis which should be approximately 0 and 3 for normally distributed random variables. 

The normality assumptions proposed in the method section matter for the noise estimates but the EB methodology ensures that the poverty estimates are unbiased. The residual analysis suggests that the skewness and kurtosis of the idiosyncratic and district level area effects match the normality assumptions. However, there appears to be few outliers within the error term normal density plot \@ref(tab:tab3). The residual plots for both the random error and idiosyncratic errors can be found below: 


```{r fig1, echo = FALSE, fig.cap = "Fay-Herriot Residual Plots"}

### density plots for the idiosyncratic errors and random effects


plot_random <- 
  fhmodel_not$model$random_effects[,1] %>%
  as.data.table() %>%
  setnames(new = "values") %>%
  ggplot() + 
  geom_density(aes(x = values), fill = "blue", alpha = 0.5) + 
  xlim(-0.3, 0.3) + 
  labs(x = "Random Effects", y = "Density", title = "Random Effect Residuals (\u03BC)") + 
  theme_minimal()

plot_error <- 
  unname(fhmodel_not$model$std_real_residuals[,1]) %>%
  as.data.table() %>%
  setnames(new = "values") %>%
  ggplot() + 
  geom_density(aes(x = values), fill = "blue", alpha = 0.5) + 
  xlim(-2.5, 2.75) + 
  labs(x = "Error Terms", y = "Density", title = "Standardized Error Residuals (\u03B5)") + 
  theme_minimal()

grid.arrange(grobs = list(plot_random, plot_error), nrow = 1)

xx <- shapiro.test(fhmodel_not$model$std_real_residuals)

yy <- shapiro.test(fhmodel_not$model$random_effects)

```

We employ the Shapiro-Wilks measure to test the null hypothesis that random variables, $e_i$ and $\mu_i$, come from a normality distributed populations. The test results, $W_{e_i} =$ `r specify_decimal(xx$statistic[[1]], 3)` ($p =$ `r specify_decimal(xx$p.value[[1]], 3)`); 
$W_{\mu_i} =$ `r specify_decimal(yy$statistic[[1]], 3)` ($p =$ `r specify_decimal(yy$p.value[[1]], 3)`) suggest normally distributed random effects and idiosyncratic error terms. We cannot reject the null hypothesis at the 5% level (although the standardized error residuals are significant at 10%).  

The Fay Herriot model employs direct estimates in predicting poverty rates. The shrinkage factor measures the ratio of the random effect to the total variance within the model. Full shrinkage $\hat\gamma_i$ = 1 means predicted poverty rates are simply the direct estimates while the other extreme uses a purely synthetic predictions, $\hat\gamma_i$ = 0. We present a scatter plot of the $\hat\gamma_i$ as a function of sample size. 

```{r fig1b, echo = FALSE, fig.cap = "Scatter Plot of Shrinkage Factor vs Sample Size"}

cv_dt %>%
  merge(fhmodel_not$model$gamma) %>%
  ggplot() + 
  geom_point(aes(x = survey_size, y = Gamma)) + 
  labs(
    y = expression(paste("shrinkage factor, (",hat(gamma)[i], ")")),
    x = "Sample Size"
  ) + 
  theme_bw()

```

In a final check, we attempt to validate the model by perform the Remove-One Model validation. Since our sample only contains 48 target areas, the typical n-fold validation process would have to split an already limited sample into 2 smaller training and test sets. Instead, the Remove-One validation process, trains a Fay Herriot model on 47 districts and removes 1 until every district has been excluded once. We compare show a plot comparing model validation estimates with the actual FH model predictions to check the stability of the model. 

```{r fig1c, echo = FALSE, fig.cap = "Remove-One Model Valiadation Plot"}

# Fit the linear model
lm_model <- lm(FH_validation ~ FH, data = result_dt)
correlation <- summary(lm_model)$r.squared ^ 0.5

# Create the plot
ggplot(result_dt, aes(x = FH, y = FH_validation)) +
  geom_point(color = 'blue') +                             # Scatter plot
  geom_abline(slope = 1, intercept = 0,                    # 45-degree line
              color = 'red', linetype = "dashed") +
  theme_minimal() +                                        # Minimal theme for a cleaner look
  labs(title = "Remove-one Model Validation",
       x = "Remove-one validation estimates",
       y = "FH model estimates") +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  # Add the R-squared as text annotation
  annotate("text", x = 0.1, y = 0.9, 
           label = paste0("correlation = ", round(correlation, 3)),
           color = "black", hjust = 0)

```

\@ref(fig:fig1c) plots the FH model estimates against their corresponding estimates as a result of the remove-one validation model. The correlation between the set of FH model estimates and the validation estimates stands at 0.75 as shown in above chart. 

### Poverty Maps
As a final check, the FH poverty rates at the district level are aggregated to the regional level to compare against the direct estiamtes. The regional level is the highest level of resolution at which survey design specifies national representativeness. The direct estimates in \@ref(fig:fig2) are shown as 95% confidence intervals (in red) which are plotted in comparison with Fay Herriot poverty estimates. 

```{r fig2, echo = FALSE, fig.cap = "Province Fay-Herriot Estimates"}

provpov_dt %>%
  ggplot(aes(x = admin1Name)) +
  geom_point(aes(y = provFH), color = "blue", size = 2) +  # Plotting provFH
  geom_errorbar(aes(ymin = DirectLB, ymax = DirectUB), width = 0.2, color = "red") +  # Error bars for DirectLB and DirectUB
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  # Adding a horizontal line at y = 0
  labs(x = "Province", 
       y = "Poverty Rate",
       caption = "Note: Author calculations of Direct Estimate 95% CI at the \nprovince level (Red error bars) compared with FH
       estimates at same level (blue dots)") +  # Labeling axes
  theme_bw() +  # Setting a white background theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Angling x-axis text
        plot.caption = element_text(hjust = 0.5, size = 10)) 

```
It should be noted that all the model based regional estimates appear to be lie within the direct estimates confidence intervals in all the regions.  

The poverty map corresponding to the 2022 for Somalia at the district level in the subsequent figures: 

```{r fig3, echo = FALSE, fig.cap = "Somalia Head Count Maps (Rate & Population)"}

### maps for population of poor and head count rates
povshp_dt %>%
  ggplot() +
  geom_sf(aes(fill = FH)) +
  scale_fill_viridis(option = "H") +
  theme_bw() +
  labs(fill = "Head Count \nPoverty Rate")

povshp_dt %>%
  ggplot() +
  geom_sf(aes(fill = FH * estimated_population_current)) +
  scale_fill_viridis(option = "H") +
  theme_bw() +
  labs(fill = "Population \n of Poor")

```



Upon review of the initial estimates in collaboration with SNBS, 3 districts were identified as having unrealistically low estimates of poverty as the surrounding districts within each region had significantly higher poverty rates. These districts were Garbahaarey, Lassqoray and Zeylac in Gedo, Sanaag and Awdal regions respectively. Three benchmarking approaches were implemented in attempts to solve the problem:  

i) First, the raking benchmarking method iteratively adjusts district estimates until convergence is reached with the regional poverty rate. However, the FH model regional poverty rates are all within 5% percent of the direct estimates, as a result this had little effect in changing the district poverty rates. 

ii) Next, the ratio method adjusts the district estimates using a constant factor 

iii) Finally, a method which incorporates the MSE estimates was also applied.  

All three methods had minimal effect on the district poverty rates as they all sensitive to the accuracy of sampling in the specific districts. The decision was made to treat all three districts as out of sample, which resulted in poverty rates more aligned with the neighboring districts. The results of this exercise can be found in the supplementary section. 

### Appendix

```{r fig4, echo = FALSE, fig.cap = "Correlation between FH Model Estimates and Direct Estimates at District Level"}

result_dt[!is.na(Direct),] %>%
  ggplot() +
  geom_point(aes(x = Direct,
                 y = FH)) +
  geom_abline(intercept = 0,
              slope = 1,
              color = "red") +
  ylab("FH Model Estimates") +
  xlab("Direct Estimates") +
  theme_bw()

```


```{r fig5, echo = FALSE, fig.cap = "Direct survey estimates of headcount poverty"}

provpov_dt %>%
  dplyr::select(admin1Name, Direct, DirectLB, DirectUB) %>%
  mutate(Direct = specify_decimal(Direct, 3)) %>%
  mutate(DirectLB = specify_decimal(DirectLB, 3)) %>%
  mutate(DirectUB = specify_decimal(DirectUB, 3)) %>%
  flextable() %>%
  colformat_num(j = c("Direct", "DirectLB", "DirectUB"), digits = 3) %>%
  set_header_labels(admin1Name = "Province",
                    Direct = "Direct Estimate",
                    DirectLB = "Lower Bound",
                    DirectUB = "Upper Bound") %>%
  add_header_row(
    values = c("", "Survey", "FH Estimate 95% Confidence Intervals"),
    colwidths = c(1, 1, 2)
  ) %>%
  autofit() %>%
  merge_h(part = "header") %>%
  align(align = "center", part = "header") %>%
  bg(i = ~Direct < DirectLB | Direct > DirectUB, bg = "red") %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  theme_box()

```

```{r, echo = FALSE}

#### include poverty map in table format
result_dt %>%
  merge(provpov_dt[, c("admin1Name", "admin1Pcod"), with = FALSE],
        by = "admin1Pcod") %>%
  dplyr::select(admin1Name, admin2Name, Direct, FH) %>%
  mutate(Direct = specify_decimal(Direct, 3)) %>%
  mutate(FH = specify_decimal(FH, 3)) %>%
  as.data.table() %>%
  setnames(old = c("admin1Name", "admin2Name", "Direct", "FH"),
           new = c("Region", "District", "Direct Estimate", "FH Model Estimate")) %>%
  flextable() %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() 


```



### References





