---
title: "Small Area Estimation of Poverty in Somalia : A Fay Herriot Model Approach"
author: "Ifeanyi Edochie"
date: "2024-06-20"
output: 
  bookdown::word_document2:
    reference_docx: report_style.docx
fontsize: 10pt
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

devtools::load_all()
# BiocManager::install("kable", update = FALSE, validate = FALSE)

pacman::p_load(here, ggplot2, sf, sfnetworks, kableExtra, dplyr, data.table, 
               viridis, paletteer, ggthemes, gridExtra, grid, flextable,
               showtext, extrafont, bookdown, officer, ggExtra)

```

Supplementary Section: Re-estimating FH Model without Garbahaarey, Lassqoray and Zeylac in SIHBS sample

The National Statistical Office in Somalia flagged the aforementioned areas as possessing abnormally low poverty rates. Upon further inspection, Garbahaarey and Zeylac have only 2 enumeration areas. Table S\@ref(suptab:subtab1) presents the fewest 10 districts in number of EAs sampled. 

```{r, echo = FALSE}

smoothcheck_dt <- read.csv("data-clean/ebp_results/smoothcheck.csv")
smoothcheck_dt <- smoothcheck_dt[,-1]

### compute the areas with the highest 

smoothcheck_dt[order(smoothcheck_dt$num_ea),] %>%
  dplyr::arrange(num_ea) %>%    # Order by num_ea
  head(n = 10) %>%
  flextable() %>%
  set_header_labels(admin2Pcod = "District Code",
                    admin2Name = "District Name",
                    num_hhs = "Number\nof Households",
                    num_ea = "Number\nof EAs") %>%
  bg(i = ~ admin2Name %in% c("Garbahaarey", "Lassqoray", "Zeylac"), bg = "#FFDDC1", part = "body") %>%
  add_footer_lines("The colored districts were flagged by the NSO") %>%
  align(align = "center", part = "header") %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() %>%
  theme_box() %>%
  set_caption(caption = "Sparsely Sampled Districts",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "suptab", 
                                    bkm = "suptab1")) 



```

Finally, we remove the 3 districts flagged by the SNBS and re-iterate the entire modelling exercise previously described including both the model/variable selection process and the FH model estimation. The results are as follows:

```{r, echo = FALSE}

fh_model_notrans <- readRDS("data-clean/ebp_results/fh_model_notrans_raw_sup.RDS")

reg_table <- 
fh_reportcoef_table(fh_model_notrans) %>%
    setnames(new = c("Variables", "Coefficients", "Standard Error")) %>%
    mutate(Variables = case_when(
    Variables == "(Intercept)" ~ "Intercept",
    Variables == "prodmt_reg_shr_trof" ~ "Production share in Tropical Fruit Production",
    Variables == "share_pop_in_2km_grid" ~ "Average share of population within 2km grids",
    TRUE ~ Variables
  )) %>%
  flextable() %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() %>%
  set_caption(caption = "EBP Model (Regression Results)",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab2")) %>%
  theme_box()

# # Create individual summary tables as flextables
# rsq_table <- data.table(indicator = "Model R\u00B2", 
#                         Marginal = specify_decimal(fh_normality_fit(fh_model_notrans)[1,2], 3),
#                         Conditional = specify_decimal(fh_normality_fit(fh_model_notrans)[2,2], 3)) %>%
#   flextable() %>%
#   set_header_labels(indicator = "", Marginal = "Marginal R²", Conditional = "Conditional R²") %>%
#   autofit() %>%
#   theme_box()
# 
# eps_table <- data.table(indicator = "(Error Term) \u03B5", 
#                         Skewness = specify_decimal(fh_normality_fit(fh_model_notrans)[3,2], 3),
#                         Kurtosis = specify_decimal(fh_normality_fit(fh_model_notrans)[4,2], 3)) %>%
#   flextable() %>%
#   set_header_labels(indicator = "", Skewness = "Skewness", Kurtosis = "Kurtosis") %>%
#   autofit() %>%
#   theme_box()
# 
# rdm_table <- data.table(indicator = "(Random Effect) \u03BC", 
#                         Skewness = specify_decimal(fh_normality_fit(fh_model_notrans)[5,2], 3),
#                         Kurtosis = specify_decimal(fh_normality_fit(fh_model_notrans)[6,2], 3)) %>%
#   flextable() %>%
#   set_header_labels(indicator = "", Skewness = "Skewness", Kurtosis = "Kurtosis") %>%
#   autofit() %>%
#   theme_box()


#### present the tables
reg_table


```


```{r, echo = FALSE}

norm_dt <- 
fh_normality_fit(fh_model_notrans) %>%
  mutate(value = specify_decimal(value, 3)) %>%
  t() %>%
  as.data.frame() 

colnames(norm_dt) <- norm_dt[1,]
norm_dt <- norm_dt[-1,]

norm_dt %>%
  as.data.frame() %>%
  flextable() %>%
  set_header_labels(rsq_marginal = "marginal",
                    rsq_conditional = "conditional",
                    epsilon_skewness = "skewness",
                    epsilon_kurtosis = "kurtosis",
                    random_skewness = "skewness",
                    random_kurtosis = "kurtosis") %>%
  add_header_row(values = c("Model R\u00B2", "(Error Term) \u03B5", "(Random Effect) \u03BC"), 
                 colwidths = c(2, 2, 2)) %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  theme_box() %>%
  autofit() %>%
  set_caption(caption = "Assessing Normality Assumptions",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab3"))

```

```{r, echo = FALSE, include = FALSE, warnings = FALSE, message=FALSE}

source(here("analysis/02_fayharriotmodel-supplementary.R"))

```



```{r figs1, echo=FALSE, fig.show='hold', fig.cap="Poverty Maps (Population Poor & Rate)"}

povshp_dt %>%
  ggplot() +
  geom_sf(aes(fill = FH)) +
  scale_fill_viridis(option = "H") +
  theme_bw() +
  labs(fill = "Poverty Rate")

# ggsave("figures/poverty_map_fhmodelnot_sup.png")

povshp_dt %>%
  ggplot() +
  geom_sf(aes(fill = FH * estimated_population_current)) +
  scale_fill_viridis(option = "H") +
  theme_bw() +
  labs(fill = "Population \nof Poor")

# ggsave("figures/poverty_map_count_fhmodelnot_sup.png")

```


```{r fig4, echo = FALSE, fig.cap = "Correlation between FH Model Estimates and Direct Estimates at District Level"}

result_dt[!is.na(Direct),] %>%
  ggplot() +
  geom_point(aes(x = Direct,
                 y = FH)) +
  geom_abline(intercept = 0,
              slope = 1,
              color = "red") +
  ylab("FH Model Estimates") +
  xlab("Direct Estimates") +
  theme_bw()

```


```{r, echo = FALSE}

provpov_dt %>%
  dplyr::select(admin1Name, Direct, DirectLB, DirectUB) %>%
  mutate(Direct = specify_decimal(Direct, 3)) %>%
  mutate(DirectLB = specify_decimal(DirectLB, 3)) %>%
  mutate(DirectUB = specify_decimal(DirectUB, 3)) %>%
  flextable() %>%
  colformat_num(j = c("Direct", "DirectLB", "DirectUB"), digits = 3) %>%
  set_header_labels(admin1Name = "Province",
                    Direct = "Direct Estimate",
                    DirectLB = "Lower Bound",
                    DirectUB = "Upper Bound") %>%
  add_header_row(
    values = c("", "Survey", "FH Estimate 95% Confidence Intervals"),
    colwidths = c(1, 1, 2)
  ) %>%
  autofit() %>%
  merge_h(part = "header") %>%
  align(align = "center", part = "header") %>%
  bg(i = ~Direct < DirectLB | Direct > DirectUB, bg = "red") %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  theme_box() %>%
  font(fontname = "Times New Roman", part = "all") %>%
  theme_box() %>%
  set_caption(caption = "Comparing FH Estimates to Direct Estimates at Regional Level",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab6"))

```


```{r, echo = FALSE}

original_dt <- read.csv("data-clean/ebp_results/povertymap_som.csv")

original_dt <- original_dt %>%
  rename(previousFH = FH)

#### include poverty map in table format
result_dt %>%
  merge(provpov_dt[, c("admin1Name", "admin1Pcod"), with = FALSE],
        by = "admin1Pcod", all.x = TRUE) %>%
  merge(original_dt[, c("targetarea_codes", "previousFH")], 
        all.x = TRUE, by.x = "Domain", by.y = "targetarea_codes") %>%
  dplyr::select(admin1Name, admin2Name, Direct, FH, previousFH) %>%
  mutate(Direct = specify_decimal(Direct, 3)) %>%
  mutate(FH = specify_decimal(FH, 3)) %>%
  as.data.table() %>%
  setnames(old = c("admin1Name", "admin2Name", "FH", "previousFH"),
           new = c("Region", "District", "FH Model Estimate", 
                   "Original FH Model Estimate")) %>%
  flextable() %>%
  bg(i = ~ District %in% c("Garbahaarey", "Laasqoray", "Zeylac"), bg = "#FFDDC1", part = "body") %>%
  set_table_properties(width = 0.5, layout = "autofit") %>%
  fontsize(size = 10) %>%
  theme_box() %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit() %>%
  set_caption(caption = "District-Level Poverty Map Table",
              style = "Table Caption",
              autonum = run_autonum(seq_id = "tab", bkm = "tab7"))



```



```{r, echo = FALSE}

provpov_dt %>%
  ggplot(aes(x = admin1Name)) +
  geom_point(aes(y = provFH), color = "blue", size = 2) +  # Plotting provFH
  geom_errorbar(aes(ymin = DirectLB, ymax = DirectUB), width = 0.2, color = "red") +  # Error bars for DirectLB and DirectUB
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  # Adding a horizontal line at y = 0
  labs(x = "Province", 
       y = "Poverty Rate",
       caption = "Note: Author calculations of Direct Estimate 95% CI at the \nprovince level (Red error bars) compared with FH
       estimates at same level (blue dots)") +  # Labeling axes
  theme_bw() +  # Setting a white background theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Angling x-axis text
        plot.caption = element_text(hjust = 0.5, size = 10)) 


```

































